\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-05-10 paper.tex
update convergence criterion

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

@@

\paragraph*{The QNMCEM algorithm.}
Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}. The $Condition$ in algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w)})$ is coefficient of variation at the $(w + 1)$-th iteration computed as
\begin{equation*}
  \label{eq:coefficient_variation}
  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
\end{equation*}
where sd(.) and mean(.) are the sample standard deviation and mean functions respectively and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
\begin{equation*}
  \label{eq:relative_difference}
  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}
\end{equation*}

\begin{algorithm}[]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

>>>>

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{equation}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{equation}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

@@

\paragraph*{The QNMCEM algorithm.}
Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.

% \vspace{.5cm}

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
The convergence condition in Algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w+1)})$ represents the variation at the $(w + 1)$-th iteration computed as
\begin{equation*}
  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
\end{equation*}
where sd(.) and mean(.) are the empirical standard deviation and mean functions respectively, and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
\begin{equation*}
  \label{eq:relative_difference}
  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}.
\end{equation*}
